---
title: "Reproducible Research - Peer Assessment 1"
author: "John James"
date: "March 6, 2016"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = 'C:/Users/John/Documents/GitHub/RepData_PeerAssessment1/Code/Final Scripts')
```

## Introduction

It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the “quantified self” movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks.  

## Research Question
This analysis seeks to address the following research question:

> What activity and behavior patterns can be ascertained from an a retrospective observation of movement data obtained from these enthusiasts?

More specifically, 

1. What is mean total number of steps taken per day?
2. What is the average daily activity pattern?
3. How do these observations change with imputed data?
4. Are there differences in activity patterns between weekdays and weekends?

## Data
This project makes use of data from a personal activity monitoring device. This device collects data at # 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

The data for this assignment was downloaded from the course web site: https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip

The variables included in this dataset are:

* steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)
* date: The date on which the measurement was taken in YYYY-MM-DD format
* interval: Identifier for the 5-minute interval in which measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

## The Analysis
The has been organized as follows. 

1.  Load Data - Load the raw data into the environment 
2.  Mean Average Total - Calculates mean average totalsteps per day and produces a histogram
3.  Activity Pattern - Prepares a time series of activity, in terms of steps per 5 minute interval across all days
4.  Impute Values - Missing values are replaced by the average number of steps per day across all days for the missing time interval
5.  Comparison - Compared density, histogram and box plots for the raw and processed data are prepared 
6.  Weekend/Weekday - Time series line charts are prepared for both weekday and weekends.

### Load Data

```{r echo = F}
library(knitr)
read_chunk("load.R")
read_chunk("process.R")
read_chunk("func.R")
```

```{r echo = FALSE, message =F, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE)}
<<initialize>>
<<housekeeping>>
```

The following code downloads the data from the course website and loads it into a data frame for processing.  To avoid overloading the server, there is functionality to ensure that the raw data is only downloaded once per day.
```{r echo = TRUE, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE)}
<<loadDataFunc>>
```

```{r echo = TRUE, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE)}
<<loadDataCall>>  
```

### Mean Average Total
The following functionality explores the average total number of steps taken per day. It produces a histogram with summary statistics showing the frequency distribution of total average steps per day.

First, we obtain only the complete cases from the raw data set.
```{r echo = TRUE, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE)}
<<completeCases>>
```

Next, the summate function summarizes the raw data by average total steps taken per day.   
```{r echo = TRUE, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE)}
<<summateFunc>>
```

```{r echo = TRUE, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE)}
<<stepsByDay01Call>>  
```

Compute the summary statistics for average total steps per day and place that into a table that will be processed and rendered onto the histogram.
```{r echo = TRUE, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE)}
<<computeStatsFunc>>
```

```{r echo = TRUE, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE)}
<<computeStats01Call>>  
```
Prepare and render the histogram with summary statistics.

```{r echo = TRUE, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE)}
<<makeHistFunc>>
```

```{r echo = TRUE, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE)}
<<makeHist01Call>>  
```

The total average number of steps per day has a range of approximately 20,000 steps, with most of the values concentrated between 10,000 and 15,000 steps per day.  The remarkably high density of observations at the median/mean suggests a homogeneity within the sample group, especially regarding the amount of daily activity. 

### Daily Activity Pattern
This section of the analysis seeks to illuminate patterns in the daily activity of the subjects.  A time series plot of the 5-minute interval and the average number of steps taken, averaged across all days (y-axis) is presented showing the 5-minute interval, on average across all the days in the dataset, that contains the maximum number of steps?

Using the summate function above, we summarize the observations by average number of steps per 5-minute interval.  
```{r echo = FALSE, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE)}
<<summateFunc>>
```

```{r echo = TRUE, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE)}
<<stepsByInterval01>>  
```

Next we call the makeTimeSeriesFunc to produce the line graph.

```{r echo = TRUE, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE)}
<<makeTimeSeriesFunc>>
```

```{r echo = TRUE, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE)}
<<makeTimeSeriesCall>>  
```

It would appear that the most active 5 minutes of the day are between 8:35 and 8:40 am.

### Impute Values
This section of the analysis seeks to illuminate patterns in the daily activity of the subjects.  A time series plot of the 5-minute interval and the average number of steps taken, averaged across all days (y-axis) is presented showing the 5-minute interval, on average across all the days in the dataset, which contains the maximum number of steps?

First we ascertain the number of NA values in the raw dataset.
```{r echo = TRUE, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE)}
<<missingValues>>  
```

Next, we impute the data.  The strategy was to replace the missing value for each 5 minute interval with the average for that 5 minute interval calculated over all the days of the study.
```{r echo = TRUE, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE)}
<<imputeFunc>>
```

```{r echo = TRUE, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE)}
<<imputeCall>>  
```

Next the processed data is summarized by average total steps taken each day.  The summate function is used for this calculation.
```{r echo = FALSE, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE)}
<<summate>>
```

```{r echo = TRUE, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE)}
<<stepsByDay02>>  
```

Next, summary statistics are calculated to show the mean and median average total steps per day.
```{r echo = TRUE, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE)}
<<computeStatsFunc>>
```

```{r echo = TRUE, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE)}
<<computeStatsCall>>  
```

Finally, the histogram and summary statistics are prepared and rendered using the makeHist function from above.
```{r echo = FALSE, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE)}
<<makeHistFunc>>
```

```{r echo = TRUE, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE)}
<<stepsHist02>>  
```

### Comparison
At this stage, we seek to illustrate the ways in which these values differ from the estimates from the first part of the assignment and the impact of imputing missing data on the estimates of the total daily number of steps.

First we conjoin the raw and processed data into a single data frame.
```{r echo = TRUE, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE)}
<<conjoinFunc>>
```

```{r echo = TRUE, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE)}
<<conjoinCall>>  
```


```{r echo = TRUE, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE)}
<<makeDensityPlotFunc>>
```

```{r echo = TRUE, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE)}
<<makeDensityPlotCall>>  
```

We observe that their is a higher density of average total steps per day near the median. This is expected, given the strategy we used to impute the data.

The histogram confirms a significantly higher concentration of values at the mean.
```{r echo = TRUE, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE)}
<<makeOverlaidHistFunc>>
```

```{r echo = TRUE, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE)}
<<makeOverlaidHistCall>>  
```

Next we compute the difference in summary statistics for each data set.
```{r echo = TRUE, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE)}
<<compareStatsFunc>>
```

```{r echo = TRUE, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE)}
<<compareStatsCall>>  
```

Finally, the boxplot illuminates the differences in the summary statistics.
```{r echo = TRUE, messages=FALSE, warnings=FALSE, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE),fig.height=12}
<<makeBoxPlotFunc>>
```

```{r echo = TRUE, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE)}
<<makeBoxPlotCall>>  
```

There was an 11% increase in average total steps per day in the 25th percentile; however, the means and medians were remarkly close within and across data sets.  Notwithstanding, imputing the data using our scheme had little affect on the estimates.

### Weekend/Weekday Activity Analysis
Our final step in this analysis is to observe whether there was any significant difference in activity during the weekdays vis-a-vis weekends.  For this analysis, we show comparative 5-minute interval time series for both weekends and weekdays.

First, we add a categorical factor value to the data set that indicates whether an observation was on a weekday or a weekend.
```{r echo = TRUE, messages=FALSE, warnings=FALSE, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE),fig.height=12}
<<dayLabelFunc>>
```

```{r echo = TRUE, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE)}
<<dayLabelCAll>>  
```

Next we summarize average total steps per day, by the weekday/weekend categorical factor.
```{r echo = TRUE, messages=FALSE, warnings=FALSE, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE),fig.height=12}
<<summateWdFunc>>
```

```{r echo = TRUE, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE)}
<<summateWdCall>>  
```

Lastly, we produce the time series.
```{r echo = TRUE, messages=FALSE, warnings=FALSE, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE),fig.height=12}
<<makeTimeSeriesWdFunc>>
```

```{r echo = TRUE, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE)}
<<makeTimeSeriesWdCall>>  
```

As one might expect, the data shows less activity during the morning hours on the weekends, but significantly greater activity starting during the early afternoon and into the early evening, then peaking once again around 2100 hours.